---
title: "Effect of Walking Fitness on Brain Health"
author: "Edward Huffstetler"
output: html_document
date: "2026-03-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

Github repository contains summary PDF (results only/no code), and instructions for how to run pipeline with real data:
https://github.com/Edward932/Brain-Age-Gap-Demo-Project

This markdown file is the reproducible demo pipeline for the Brain Age Gap (BAG) analysis summarized in Bag_Summary.PDF. It trains an Elastic Net and XGBoost model to predict age from \~1,400 MRI-derived summary measures. It then computes BAG (predicted age - true age) and tests whether two-minute walk distance is associated with BAG.

Because the raw AABC data are governed by a data-use agreement, this Rmd runs by default using synthetic data that mimics the structure of the real dataset. To use the real data follow the instructions in the README. If you use synthetic data (the default option) the charts and graphs will not match BAG_Summary.PDF.

# Setup

This notebook assumes packages are installed. If you cloned the repo, the recommended way to install the same package versions is:

install.packages("renv"); renv::restore()

```{r load packages and set seed, echo=FALSE, message=FALSE, warning=FALSE}
required = c("dplyr","knitr","glmnet","ggplot2","xgboost",
              "mlr3","mlr3mbo","mlr3learners")

missing = required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing packages: ", paste(missing, collapse = ", "),
       "\nIf you cloned the repo, run: install.packages('renv'); renv::restore()",
       "\nOtherwise install packages with: install.packages(c(",
       paste0('"', missing, '"', collapse = ", "), "))")
}

library(dplyr)
library(knitr)
library(glmnet)
library(ggplot2)
library(xgboost)
library(mlr3)
library(mlr3mbo)
library(mlr3learners)
```

# Data Loading

Synthetic data is generated by default. To use real data follow the instructions in the README (requires downloading data).

```{r load data}
# To save time some models are pre-trained and loaded in
# If you change the seed, the pre-trained models may not reflect the new synthetic data
set.seed(1)

# Importing functions
source("DataProcessing.R")

# By default pipeline uses synthetic data (real data requires DUA access)
# Instructions to use real data are in the README
# Changing to FALSE will only work if instructions in README are followed
USE_SYNTHETIC_DATA = TRUE
if (USE_SYNTHETIC_DATA) {
  df_list = generate_synthetic_data()
} else {
  df_list = clean_real_data()
}

# MRI_df contains MRI summary measurements and age
MRI_df = df_list$MRI
```

# Model 1: Elastic Net

First, we fit an Elastic Net model to predict chronological age from MRI summary measurements. Predictions are generated out-of-fold via cross-validation to avoid leakage. These out-of-fold predictions are then used to compute BAG.

Because brain-age models often show age-dependent bias, we also apply a simple linear calibration of predicted age on true age using out-of-fold predictions, and then compute calibrated BAG. This calibration is intended to remove age dependence in BAG and may not improve RMSE.

For downstream analysis BAG is calculated for each patient, so data is not split into training and testing. Instead out-of-fold predictions are calculated for each patient.

```{r elastic net}
# Separating data into age (y_ridge) and MRI measurements (x_ridge) for training a elastic net model to predict age based on MRI data
y_ridge = MRI_df$age
x_ridge = as.matrix(MRI_df[ , !(names(MRI_df) %in% c("ID", "age"))])

# Training a elastic net model using 10 fold cross validation
# Out-of-fold age predictions are saved to calculate BAG
# Alpha of 0.014 was selected via brute force method
elastic_net_model = cv.glmnet(x_ridge, y_ridge, alpha = 0.014, nfolds = 10, standardize = TRUE, keep = TRUE)

# Extracting out-of-fold predictions for the optimal lambda value determined through CV
best_fit_index = elastic_net_model$index["min", "Lambda"]
pred_oof_EN = as.numeric(elastic_net_model$fit.preval[, best_fit_index])

# calculating BAG for each patient and root mean squared error of age prediction (no calibration)
bag_EN = pred_oof_EN - y_ridge
RMSE_EN = sqrt(mean((pred_oof_EN - y_ridge)^2))

# Using linear calibration to correct age-dependent bias in predictions
# Then rescaling predictions so the fitted line is y = x
cal_EN = lm(pred_oof_EN ~ y_ridge)
intercept_EN = coef(cal_EN)[1]
slope_EN = coef(cal_EN)[2]
pred_oof_corr_EN = (pred_oof_EN - intercept_EN) / slope_EN

# calibrated predicted age and calibrated BAG
bag_corr_EN = pred_oof_corr_EN - y_ridge
RMSE_corr_EN = sqrt(mean((pred_oof_corr_EN - y_ridge)^2))
```

# Elastic Net prediction plots before and after calibration

```{r elastic net graphs, echo = FALSE}
# Formating data for graph
label = paste("Linear fit\ny ≈", signif(coef(cal_EN)[[1]], 5), "+", signif(coef(cal_EN)[[2]], 3), "x\nRMSE", signif(RMSE_EN, 5))
df_EN = data.frame(y_true = y_ridge, y_pred = pred_oof_EN)

ggplot(df_EN, mapping = aes(x = y_true, y = y_pred)) +
  geom_point(size=1, alpha = 0.15, position = position_jitter(width = 0.25, height = 0)) +
  geom_abline(aes(color = label, intercept = coef(cal_EN)[1], slope = coef(cal_EN)[[2]]), linewidth = 0.8) +
  geom_abline(aes(color = "Identity (y = x)", intercept = 0, slope = 1), linewidth = 0.8, linetype = "dashed") +
  labs(title = "True Age vs Predicted Age (EN)", x = "True Age (chronological)", y = "Predicted Age", subtitle="Out-of-fold age predicton from MRI") +
  scale_color_manual(values = c("Identity (y = x)" = "black", setNames("blue", label))) +
  labs(color = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2), legend.background = element_rect(fill="white", color=NA), legend.key = element_rect(fill="white", color=NA)) +
  coord_cartesian(ylim = c(20, 110))

# Formating data for graph
label_corr = paste("Identity (y = x)\nRMSE", signif(RMSE_corr_EN, 5))
df_corr_EN = data.frame(y_true = y_ridge, y_pred = pred_oof_corr_EN)

ggplot(df_corr_EN, mapping = aes(x = y_true, y = y_pred)) +
  geom_point(size=1, alpha = 0.15, position = position_jitter(width = 0.25, height = 0)) +
  geom_abline(aes(color = label_corr, intercept = 0, slope = 1), linewidth = 0.8, linetype = "dashed") +
  labs(title = "True Age vs Predicted Age (EN)", x = "True Age (chronological)", y = "Predicted Age (calibrated)", subtitle="Out-of-fold predicton bias corrected by linear calibration") +
  scale_color_manual(values = setNames("black", label_corr)) +
  labs(color = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2), legend.background = element_rect(fill="white", color=NA), legend.key = element_rect(fill="white", color=NA)) +
  coord_cartesian(ylim = c(20, 110))
```

# Model 2: XGBoost

Next, we train an XGBoost model using cross-validated out-of-fold predictions. Hyperparameters were selected using a model-based optimization procedure (mlr3mbo) over a predefined search space, minimizing cross-validated RMSE. Hyperparameter optimization code is in the appendix at the bottom of this Rmd.

To keep knitting time manageable, the notebook can load cached cross-validation results in FAST_MODE. If FAST_MODE = FALSE, the full CV is recomputed (this may take a few minutes).

```{r xgboost}
# Separating data into age (y_xgb) and MRI measurements (x_xgb) for training a xgboost model to predict age based on MRI data
y_xgb = MRI_df$age
x_xgb = as.matrix(MRI_df[ , !(names(MRI_df) %in% c("ID", "age"))])

# Converting into XGBoost’s optimized DMatrix format
xgb_matrix = xgb.DMatrix(data = x_xgb, label = y_xgb)

# Creating a list of optimized hyperparameters
# hyperparameters where chosen by Bayesian-optimization-like procedure via the mlr3mbo package
# Code for hyperparameter optimization is located at the end of this file
params = list(objective = "reg:squarederror",
              eval_metric = "rmse",
              tree_method = "hist",
              max_depth = 4,
              min_child_weight = 81.07821,
              subsample = 0.7210051,
              colsample_bytree = 0.1494652,
              colsample_bynode = 0.2376824,
              eta = 0.02154221,
              gamma = 4.175431,
              alpha = 3.37655,
              lambda = 1.418648)

# Training XGBoost using 10 fold cross validation
# Out-of-fold age predictions are saved to calculate BAG
# This can take a few minutes to run so previous results are being loaded
# Change value to FALSE to run CV (may take a few minutes)
FAST_MODE = TRUE
if (FAST_MODE && USE_SYNTHETIC_DATA) {
  xgb_model = readRDS("xgb_cv.rds")
} else {
  xgb_model = xgb.cv(
    params = params,
    data = xgb_matrix,
    nrounds = 20000,
    nfold = 10,
    early_stopping_rounds = 80,
    verbose = 0,
    prediction = TRUE)
}

# Extracting out-of-fold predictions
pred_oof_xgb = xgb_model$cv_predict[[1]]
pred_oof_xgb = as.numeric(pred_oof_xgb)

# calculating BAG for each patient and root mean squared error of age prediction (no calibration)
bag_xgb = pred_oof_xgb - y_xgb
RMSE_xgb = sqrt(mean((pred_oof_xgb - y_xgb)^2))

# Using linear calibration to correct age-dependent bias in predictions
# Then rescaling predictions so the fitted line is y = x
cal_XGB = lm(pred_oof_xgb ~ y_xgb)
intercept_XGB = coef(cal_XGB)[1]
slope_XGB = coef(cal_XGB)[2]
pred_oof_corr_xgb = (pred_oof_xgb - intercept_XGB) / slope_XGB

# calibrated predicted age and calibrated BAG
bag_corr_xgb = pred_oof_corr_xgb - y_xgb
RMSE_corr_xgb = sqrt(mean((pred_oof_corr_xgb - y_xgb)^2))
```

# Elastic Net prediction plots before and after calibration

```{r xgboost graphs, echo=FALSE}
# Formating data for graph
label_xgb = paste("Linear fit\ny ≈", signif(coef(cal_XGB)[[1]], 5), "+", signif(coef(cal_XGB)[[2]], 3), "* x\nRMSE", signif(RMSE_xgb, 5))
df_xgb = data.frame(y_true = y_xgb, y_pred = pred_oof_xgb)

ggplot(df_xgb, mapping = aes(x = y_true, y = y_pred)) +
  geom_point(size=1, alpha = 0.2, position = position_jitter(width = 0.25, height = 0)) +
  geom_abline(aes(color = label_xgb, intercept = coef(cal_XGB)[1], slope = coef(cal_XGB)[[2]]), linewidth = 0.8) +
  geom_abline(aes(color = "Identity (y = x)", intercept = 0, slope = 1), linewidth = 0.8, linetype = "dashed") +
  labs(title = "True Age vs Predicted Age (XGB)", x = "True Age (chronological)", y = "Predicted Age", subtitle="Out-of-fold age predicton from brain MRI summaries") +
  scale_color_manual(values = c("Identity (y = x)" = "black", setNames("blue", label_xgb))) +
  labs(color = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2), legend.background = element_rect(fill="white", color=NA), legend.key = element_rect(fill="white", color=NA)) +
  coord_cartesian(ylim = c(20, 110))

# Formatting data for graph
df_corr_xgb = data.frame(y_true = y_xgb, y_pred = pred_oof_corr_xgb)
label_corr_xgb = paste("Identity (y = x)\nRMSE", signif(RMSE_corr_xgb, 5))

ggplot(df_corr_xgb, mapping = aes(x = y_true, y = y_pred)) +
  geom_point(size=1, alpha = 0.2, position = position_jitter(width = 0.25, height = 0)) +
  geom_abline(aes(color = label_corr_xgb, intercept = 0, slope = 1), linewidth = 0.8, linetype = "dashed") +
  labs(title = "True Age vs Predicted Age (XGB)", x = "True Age (chronological)", y = "Predicted Age (calibrated)", subtitle="Out-of-fold predicton bias corrected by linear calibration") +
  scale_color_manual(values = setNames("black", label_corr_xgb)) +
  labs(color = NULL) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2), legend.background = element_rect(fill="white", color=NA), legend.key = element_rect(fill="white", color=NA)) +
  coord_cartesian(ylim = c(20, 110))
```

# Downstream analysis

For downstream, analysis we opted to use the calibrated Elastic Net BAGs. Both models had a similar RMSE (with the real data) but Elastic Net was more consistent.

We tested whether a higher two-minute walk distance is associated with calibrated brain age gap. We fit a multivariable linear regression with calibrated BAG as the outcome and two-minute walk distance as the primary predictor, adjusting for age, sex, BMI, education, and blood pressure.

```{r regreassion analysis}
# Creating a data frame with attributes such as age, sex, BMI, etc...
# Adding calibrated/corrected BAG from elastic net predictions
attributes_df = df_list$attributes
attributes_df$bag_corr = bag_corr_EN

# Fit multivariable linear regression testing whether 2-minute walk distance is 
# associated with calibrated BAG, adjusting for other variables
bag_fit = lm(bag_corr ~ two_min_walk + 
               age + 
               sex + 
               bmi + 
               education + 
               sbp_sitting,
             data = attributes_df)

# Store coefficient table
coefs = coef(summary(bag_fit))

# Create a grid of walk distances
xgrid = seq(min(attributes_df$two_min_walk), max(attributes_df$two_min_walk), length.out = 200)

# Calculating holds for other variables at mean for numerical and mode for categorical
sex_freq = sort(table(attributes_df$sex), decreasing = TRUE)
mode_sex = rownames(sex_freq)[[1]]
edu_freq = sort(table(attributes_df$education), decreasing = TRUE)
mode_edu = rownames(edu_freq)[[1]]

partial_effect = data.frame(
  two_min_walk = xgrid,
  age = median(attributes_df$age),
  sex = mode_sex,                         
  bmi = median(attributes_df$bmi),
  education = mode_edu,
  sbp_sitting = median(attributes_df$sbp_sitting)
)

# Predicted calibrated BAG along the walk-distance grid
yhat = predict(bag_fit, newdata = partial_effect)
predicted_df = data.frame(two_min_walk = xgrid, yhat = yhat)
```

# Partial-effect curve

We plot a partial-effect curve showing the calibrated BAG across the range of two-minute walk distance while holding other covariates constant.

```{r analysis graph, echo=FALSE}
ggplot(attributes_df, mapping = aes(x = two_min_walk, y = bag_corr)) +
  geom_point(size=1, alpha = 0.15, position = position_jitter(width = 0.25, height = 0)) +
  geom_line(data = predicted_df,aes(x = two_min_walk, y = yhat, color = "Model prediction (partial effect)"), linewidth = 0.8) +
  labs(title = "Partial Effect of Two Min Walk on Brain Age Gap", x = "Two Minute Walk (feet)", y = "Calibrated BAG (years)", subtitle="Prediction holds age, BMI, and SBP at medians; sex and education at modes.", color = NULL) +
  scale_color_manual(values = c("Model prediction (partial effect)" = "blue")) +
  theme_minimal() +
  theme(legend.position = c(0.25, 0.1), legend.background = element_rect(fill="white", color=NA), legend.key = element_rect(fill="white", color=NA)) +
  coord_cartesian(ylim=c(-40, 40))
```

# Effect Summary

We summarize the estimated association as change in calibrated BAG per 100 feet walked (with 95% CI), and implied change in calibrated BAG moving from the 25th to 75th percentile of walking distance.

```{r analysis table, echo=FALSE}
# Calculating slope and confidence interval for 100 foot change in 2-minute walk
beta = coef(bag_fit)[["two_min_walk"]]
ci = confint(bag_fit)["two_min_walk", ]
beta_100 = 100 * beta
ci_100 = 100 * ci

# Calculating 25th and 75th percentile and the change in BAG associated with
# going from 25th to 75th percentile
q = quantile(attributes_df$two_min_walk, probs = c(0.25, 0.75), na.rm = TRUE)
walk_25 = q[[1]]
walk_75 = q[[2]]
delta_walk = walk_75 - walk_25
delta_bag = beta * delta_walk

df_for_table = data.frame(
  Metric = c(
    "Effect per 100 feet (years BAG / 100 ft)",
    "Two-minute walk (25th percentile, feet)",
    "Two-minute walk (75th percentile, feet)",
    "Δ Walk (75th - 25th, feet)",
    "Implied Δ BAG from 25th→75th (years)"
  ),
  Estimate = c(
    beta_100,
    walk_25,
    walk_75,
    delta_walk,
    delta_bag
  ),
  "95% CI (lower)" = c(
    ci_100[1],
    NA, NA, NA, NA
  ),
  "95% CI (upper)" = c(
    ci_100[2],
    NA, NA, NA, NA
  ),
  check.names = FALSE
)

kable(df_for_table, digits = 2, caption = "Association between 2-minute walk distance and calibrated brain age gap (BAG)")

```

# Appendix

## XGBoost hyperparameter optimization (optional)

This section reproduces the model-based hyperparameter search used to select XGBoost parameters. It is turned off by default because it can take substantial time to run.

```{r parameter optimization}
# Changing VIEW_DEMO_OF_HYPERPARAMATER_OPTIMIZATION to TRUE will run demo of hyperparamater optimization for XGBOOST
# Best performing hyperparameters are already in use in the code above

# ------------------------------------------------------------ #
# WARNING: This can take over 30 minutes on a standard laptop  #
# ------------------------------------------------------------ #
VIEW_DEMO_OF_HYPERPARAMATER_OPTIMIZATION = FALSE
if (VIEW_DEMO_OF_HYPERPARAMATER_OPTIMIZATION) {
  task = as_task_regr(MRI_df[,-1], target = "age")
  
  learner = lrn("regr.xgboost",
                booster = "gbtree",
                objective = "reg:squarederror",
                eval_metric = "rmse",
                tree_method = "hist",
                nrounds = 20000,
                early_stopping_rounds = 80,
                validate = "test",
                verbose = 0)
  
  search_space = ps(
    eta = p_dbl(0.008, 0.03),
    max_depth = p_int(2L, 4L),
    min_child_weight = p_dbl(50, 140),
    subsample = p_dbl(0.55, 0.85),
    colsample_bytree = p_dbl(0.06, 0.18),
    lambda = p_dbl(0, 20),
    alpha = p_dbl(1, 4),
    gamma = p_dbl(2, 9),
    colsample_bynode = p_dbl(0.08, 0.25))
  
  resampling = rsmp("cv", folds = 5)
  resampling$instantiate(task)
  measure = msr("regr.rmse")
  tuner = tnr("mbo")
  terminator = trm("evals", n_evals = 60)
  
  instance = TuningInstanceBatchSingleCrit$new(
    task = task,
    learner = learner,
    resampling = resampling,
    measure = measure,
    search_space = search_space,
    terminator = terminator
  )
  
  print(tuner$optimize(instance))
  print(instance$result)
}
```
## Session Info

```{r session info}
sessionInfo()
```
